Bare Demo of IEEEtran.cls\\ for IEEE Conferences

### 2016-11-22 bare_conf.tex
First draft completed

%% bare_conf.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[conference]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/pkg/ifpdf
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
\usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
\graphicspath{{../png/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
\DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/pkg/algorithms
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/pkg/algorithmicx

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/array

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/pkg/subfig

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure.
% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
% corrections already built into the system in which case a warning will
% be issued if an attempt is made to load fixltx2e.sty as it is no longer
% needed.
% The latest version and documentation can be found at:
% http://www.ctan.org/pkg/fixltx2e

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/stfloats
% Do not use the stfloats baselinefloat ability as the IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that the IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/pkg/dblfloatfix

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/url
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Bare Demo of IEEEtran.cls\\ for IEEE Conferences}

@@

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

@@

% make the title area 
\maketitle
% As a general rule, do not put math, special symbols or citations 
% in the abstract

\begin{abstract} This work is focused on the prediction of affective
states using non-obtrusive sensors. Non-obtrusive sensors are preferred in this
work because of a hypothesis that states that an obtrusive sensor would alter
the affective states of a user. Specifically, the method uses data gathered from
a keyboard and mouse dynamics plugin embedded in a tutoring system designed for
the teaching of computer programming in the Python language. The tutoring system
requires from the user to practice their programming skills in order to progress
through the course, and the user’s creativity plays an important role in the
solving of programming exercises. Also a pre-processing method is applied, which
generates feature vectors consisting of 39 features representing each
interaction of the users by recording their keyboard and mouse dynamics. 149
feature vectors were generated and used as input of four different
classification algorithms: Naïve Bayes, J-48, k-NN, and Artificial Neural
Networks. These classification models are then compared by their performances
when classifying learners into five different affective states: boredom,
frustration, distraction, relaxation and engagement. Each of the classification
models achieved an accuracy of around 75\% and kappa of around 0.45, and, in
average, J-48 was the better algorithm with an accuracy of 75.97\%. Results show
that the data gathered from the non-obtrusive sensors can successfully be used
as another input to classification models in order to predict an individual's
affective states during their interaction with a programming tutoring system.
\end{abstract}

@@

% For peer review papers, you can put extra information on the cover % page as needed: 
% \ifCLASSOPTIONpeerreview % \begin{center} \bfseries EDICS Category:  3-BBND \end{center} 
% \fi % % For peerreview papers, this IEEEtran command inserts a page break and 
% creates the second title. It will be ignored for other modes. \IEEEpeerreviewmaketitle

@@

In the general context of learning there have been some approaches for affect
recognition. The work of Kapoor y Picard \cite{kapoor2005multimodal} uses a multi modal
approach using sensory information from facial expressions and postural shifts
of the learner combined with information about the learner's activity on the
computer; the learning activity was solving a puzzle on a computer. They report
using multimodal Gaussian Process approach achieving an accuracy of over 86\%.
Sidney, et al., \cite{sidney2005integrating,} enhances the intelligent tutor system AutoTutor with
affective recognition using nonintrusive sensory information from facial
expressions, gross body movements, and conversational cues from logs.  The
subject material of that system consisted in lectures of computer literacy.

@@

@@

\section{Proposed Method} 
The goal of this work is to propose an affective
recognition method based on the sensory data provided by the keyboard and mouse
dynamics generated by a learner as he/she types a programing exercise. The main
components of the method are depicted in Figure 1. A brief explanation of the
whole process is explained next, details come later. The process starts when the
learner begins to type a program in a browser-based editor. As she types or
moves the mouse the dynamics are recorded. When the learner submits the code to
evaluation, the request includes the sensory data along with the code and
information about the session. In the server, the code is evaluated in an
external virtual machine that provides a sand box to prevent malicious or
erroneous code to halt the server. When the result is ready, it is recorded
along with the sensory data and sent to the preprocessing module. The output is
a feature vector, ready for classification. A previously trained classifier is
responsible for the classification, which outputs the predicted affective state.
The method does not consider other sensory data, but it could be integrated with
other sensory inputs in a multi-modal approach.  Each of the steps is explained
in detail next.

\subsubsection{Capturing the Keystroke and Mouse Data} 
While a student is trying
to solve an exercise, a script coded in JavaScript is running in the background,
which captures every keystroke, mouse movement and mouse button press. Each
capture of these events records a timestamp in milliseconds (using the method
getTime() of Javascript's built-in class Date) that describes when the event
occurred. If the event is a keystroke, the script captures what key was
specifically pressed, and what type of event occurred, it can be either a key-
down or a key-up event. If it is an event related to a mouse button press, the
key code of that button is recorded, as well as the type of event occurred again
key-down or key-up. Finally, if the event was a mouse movement, the mouse
coordinates inside of the web browser are recorded. The script monitors the
mouse position every 100 milliseconds, and only if the position has changed, it
records the new position. Each time a learner tries to evaluate a program, all
the data generated is sent to the server along with the code. When the result of
the execution is returned, all records are cleared and the process starts again.
There is no problem if the user leaves for a long period of time, because no
event will be triggered. If a user copies and then pastes the code from another
source, this will be recorded. There is a limitation, only the browser-editor
must be used; this could be a problem for more advanced programmers needing
specialized editors.  On the other hand a browser-based editor with the
corresponding remote execution, does not require the installation of
interpreters or compilers in learner’s computer. The code could even be written
in a mobile device or any web-enabled device.  The interface of the web-based
editor is shown in Figure 2. Programming exercises are evaluated using unit
tests; the results of the evaluation are shown to users. If all tests the
program is considered to be correct. An example execution is shown in the left
side image of Figure 2. In this example the learner was asked to write a
function to add two numbers, in Python. The source code for the Protoboard web
based learning environment including the KD and MD functions are open source and
available as Github repositories at http://git.io/vJlUV also the code for the
sandbox is in http://git.io/vJlUj.

\subsubsection{Preprocessing of the Keystroke and Mouse Data} 
The raw data
obtained from the script needs to be preprocessed to obtain a feature vector.
Basically, this pre-processing consists in measuring the delays between key-
down, key-up or mouse-move events triggered during an exercise. These events
have the dynamic shown in Figure 3; for example when typing the word ‘key’ a
user first presses the letter K and triggers the key-down event, this is
indicated with an arrow that changes the state of the key, the time of the event
is important and also recorded. Only the event data is received from the
browser, in order to generate feature vectors, the patterns and rhythm users
have when pressing consecutive keys are captured measuring the delays between
events, as it is a common practice when dealing with keystroke dynamics. In this
work, the definitions proposed by Sim y Janakiraman (2007, June) are used. Held
time (Ht) is defined as the time between a key-down and a key-up of the same
key, this would be Ht(K) in the example. Inter key time (It) is defined as the
time between two consecutive key-down events, this time could be negative is the
second key is pressed before the first is released. A sequence is defined as a
list of consecutive keystrokes. In the above example valid sequences could be
[‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
third is a trigraph. While there can be sequences of any size in a text, working
only with digraphs and trigraphs is preferred. Even if only digraphs and
trigraphs are considered, the amount found in free-text is too large and not
very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
they propose to use statistical measures to capture the patterns; the same
strategy is used in this work. The averages and standard deviations are
calculated from the delays between the events of the digraphs and trigraphs in
each program.

@@

Figure 4. Web-based editor used for code evaluation. A tutorial was developed to
obtain the necessary data using Protoboard, a web-based programming learning
environment (the latest version can be found online at
http://python.protoboard.org/). Protoboard has the functionality described in
the proposed method and shown schematically in Figure 1. Users-log in to
Protoboard using their account credentials from a popular social network
(Facebook).  The web tutorial begins with three introductory videos that explain
the fundamentals of programming in Python, and how to solve the programming
exercises in the course. What follows after these videos, are 13 programming
exercises that the students need to solve in consecutively. Exercises are of
incremental difficulty. In Figures 2 and 3, the interface of this platform is
presented. For this experiment Protoboard was configured to allow unlimited
tries to solve an exercise.

A total of 55 volunteers, with no previous experience in Python, where recruited
as a response to three announcements in a special interest group of programming
students in a social network. Learners were required to login with their
Facebook credentials, this had the advantage of not requiring the system to
maintain users’ credentials or profile information as this is kept outside in an
external service. Unfortunately this decision also kept some prospects from
volunteering; a subject reported that he normally creates temporary email
accounts to try new web services. Other requirement was that they where all
adults, this age is 18 years in Mexico where the experiment was conducted. The
requirement of the social network account stopped several prospects from
volunteering, but on the other hand gave researchers access to public
information about the subjects. All the learners were either current students or
graduates of Computer Systems Engineering. Their ages were in the range of 18 to
30 years old. Although no experience in software programming was needed, as the
web tutorial's course is of a very basic level, all the participants were
required had completed at least one course in software programming. Results from
the initial survey are shown in Figure 4.

@@

Figure 5. Web-based editor used for code evaluation.

>>>>

\documentclass[conference]{IEEEtran}

% *** CITATION PACKAGES ***
%
\usepackage{cite}

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
\usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
\graphicspath{{./img/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
\DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else

\fi

\usepackage{stfloats}
% *** PDF, URL AND HYPERLINK PACKAGES ***
%
\usepackage{url}

@@

\begin{document}

% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Bare Demo of IEEEtran.cls\\ for IEEE Conferences}

@@

@@

% make the title area 
\maketitle
\begin{abstract} This work is focused on the prediction of affective
states using non-obtrusive sensors. Non-obtrusive sensors are preferred in this
work because of a hypothesis that states that an obtrusive sensor would alter
the affective states of a user. Specifically, the method uses data gathered from
a keyboard and mouse dynamics plugin embedded in a tutoring system designed for
the teaching of computer programming in the Python language. The tutoring system
requires from the user to practice their programming skills in order to progress
through the course, and the user’s creativity plays an important role in the
solving of programming exercises. Also a pre-processing method is applied, which
generates feature vectors consisting of 39 features representing each
interaction of the users by recording their keyboard and mouse dynamics. 149
feature vectors were generated and used as input of four different
classification algorithms: Naïve Bayes, J-48, k-NN, and Artificial Neural
Networks. These classification models are then compared by their performances
when classifying learners into five different affective states: boredom,
frustration, distraction, relaxation and engagement. Each of the classification
models achieved an accuracy of around 75\% and kappa of around 0.45, and, in
average, J-48 was the better algorithm with an accuracy of 75.97\%. Results show
that the data gathered from the non-obtrusive sensors can successfully be used
as another input to classification models in order to predict an individual's
affective states during their interaction with a programming tutoring system.
\end{abstract}

@@

@@

In the general context of learning there have been some approaches for affect
recognition. The work of Kapoor y Picard \cite{kapoor2005multimodal} uses a multi modal
approach using sensory information from facial expressions and postural shifts
of the learner combined with information about the learner's activity on the
computer; the learning activity was solving a puzzle on a computer. They report
using multimodal Gaussian Process approach achieving an accuracy of over 86\%.
Sidney, et al., \cite{sidney2005integrating} enhances the intelligent tutor system AutoTutor with
affective recognition using nonintrusive sensory information from facial
expressions, gross body movements, and conversational cues from logs.  The
subject material of that system consisted in lectures of computer literacy.

@@

\begin{figure}[!t] 
\centering 
\includegraphics[width=2.5in]{KMDAffective.png} 
\caption{Code Execution and Affect Recognition Architecture.}
\label{fig_process} 
\end{figure}

@@

\section{Proposed Method} 
The goal of this work is to propose an affective
recognition method based on the sensory data provided by the keyboard and mouse
dynamics generated by a learner as he/she types a programing exercise. The main
components of the method are depicted in Figure \ref{fig_process} A brief explanation of the
whole process is explained next, details come later. The process starts when the
learner begins to type a program in a browser-based editor. As she types or
moves the mouse the dynamics are recorded. When the learner submits the code to
evaluation, the request includes the sensory data along with the code and
information about the session. In the server, the code is evaluated in an
external virtual machine that provides a sand box to prevent malicious or
erroneous code to halt the server. When the result is ready, it is recorded
along with the sensory data and sent to the preprocessing module. The output is
a feature vector, ready for classification. A previously trained classifier is
responsible for the classification, which outputs the predicted affective state.
The method does not consider other sensory data, but it could be integrated with
other sensory inputs in a multi-modal approach.  Each of the steps is explained
in detail next.

\begin{table*}[!t]
\centering
\caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
    \begin{tabular}{ | l | l || l | l | }
    \hline
    ID          & Description           & ID               & Description  \\
    \hline
    1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
    \hline
    2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
    \hline
    3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
    \hline
    4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
    \hline
    5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
    \hline
    6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
    \hline
    7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
    \hline
    8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
    \hline
    9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
    \hline
    10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
    \hline
    11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
    \hline
    12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
    \hline
    13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
    \hline
    14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
   \hline
    15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
    \hline
    16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
    \hline
    17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
    \hline
    18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
    \hline
    19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
    \hline
    20 & Standard deviation of the previous feature. &   &  \\
    \hline
  \end{tabular}
\end{table*}

\subsubsection{Capturing the Keystroke and Mouse Data} 
While a student is trying
to solve an exercise, a script coded in JavaScript is running in the background,
which captures every keystroke, mouse movement and mouse button press. Each
capture of these events records a timestamp in milliseconds (using the method
getTime() of Javascript's built-in class Date) that describes when the event
occurred. If the event is a keystroke, the script captures what key was
specifically pressed, and what type of event occurred, it can be either a key-
down or a key-up event. If it is an event related to a mouse button press, the
key code of that button is recorded, as well as the type of event occurred again
key-down or key-up. Finally, if the event was a mouse movement, the mouse
coordinates inside of the web browser are recorded. The script monitors the
mouse position every 100 milliseconds, and only if the position has changed, it
records the new position. Each time a learner tries to evaluate a program, all
the data generated is sent to the server along with the code. When the result of
the execution is returned, all records are cleared and the process starts again.
There is no problem if the user leaves for a long period of time, because no
event will be triggered. If a user copies and then pastes the code from another
source, this will be recorded. There is a limitation, only the browser-editor
must be used; this could be a problem for more advanced programmers needing
specialized editors.  On the other hand a browser-based editor with the
corresponding remote execution, does not require the installation of
interpreters or compilers in learner’s computer. The code could even be written
in a mobile device or any web-enabled device.  The interface of the web-based
editor is shown in Figure \ref{fig_editor}. Programming exercises are evaluated using unit
tests; the results of the evaluation are shown to users. If all tests the
program is considered to be correct. An example execution is shown in the left
side image of Figure \ref{fig_editor}. In this example the learner was asked to write a
function to add two numbers, in Python. The source code for the Protoboard web
based learning environment including the KD and MD functions are open source and
available as Github repositories at http://git.io/vJlUV also the code for the
sandbox is in http://git.io/vJlUj.

\begin{figure}[!t] 
\centering 
\includegraphics[width=2.5in]{editorRresult.png} 
\caption{Web-based editor used for code evaluation.}
\label{fig_editor} 
\end{figure}
\begin{figure}[b!] 
\centering 
\includegraphics[width=2.5in]{KeyDyn.png} 
\caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
\label{fig_KD} 
\end{figure}
\subsubsection{Preprocessing of the Keystroke and Mouse Data} 
The raw data
obtained from the script needs to be preprocessed to obtain a feature vector.
Basically, this pre-processing consists in measuring the delays between key-
down, key-up or mouse-move events triggered during an exercise. These events
have the dynamic shown in Figure 3; for example when typing the word ‘key’ a
user first presses the letter K and triggers the key-down event, this is
indicated with an arrow that changes the state of the key, the time of the event
is important and also recorded. Only the event data is received from the
browser, in order to generate feature vectors, the patterns and rhythm users
have when pressing consecutive keys are captured measuring the delays between
events, as it is a common practice when dealing with keystroke dynamics. In this
work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
time (Ht) is defined as the time between a key-down and a key-up of the same
key, this would be Ht(K) in the example. Inter key time (It) is defined as the
time between two consecutive key-down events, this time could be negative is the
second key is pressed before the first is released. A sequence is defined as a
list of consecutive keystrokes. In the above example valid sequences could be
[‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
third is a trigraph. While there can be sequences of any size in a text, working
only with digraphs and trigraphs is preferred. Even if only digraphs and
trigraphs are considered, the amount found in free-text is too large and not
very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
they propose to use statistical measures to capture the patterns; the same
strategy is used in this work. The averages and standard deviations are
calculated from the delays between the events of the digraphs and trigraphs in
each program.

@@

\begin{figure}[!t] 
\centering 
\includegraphics[width=2.5in]{Completed.png} 
\caption{Completed activities and exercises and by learner in ascending order.}
\label{fig_completed} 
\end{figure}

\begin{figure}[!t] 
\centering 
\includegraphics[width=2.5in]{classDist.png} 
\caption{Class distribution of emotions as reported by learners.}
\label{fig_class} 
\end{figure}

A tutorial was developed to
obtain the necessary data using Protoboard, a web-based programming learning
environment (the latest version can be found online at
http://python.protoboard.org/). Protoboard has the functionality described in
the proposed method and shown schematically in Figure 1. Users-log in to
Protoboard using their account credentials from a popular social network
(Facebook).  The web tutorial begins with three introductory videos that explain
the fundamentals of programming in Python, and how to solve the programming
exercises in the course. What follows after these videos, are 13 programming
exercises that the students need to solve in consecutively. Exercises are of
incremental difficulty. In Figures 2 and 3, the interface of this platform is
presented. For this experiment Protoboard was configured to allow unlimited
tries to solve an exercise.

A total of 55 volunteers, with no previous experience in Python, where recruited
as a response to three announcements in a special interest group of programming
students in a social network. Learners were required to login with their
Facebook credentials, this had the advantage of not requiring the system to
maintain users’ credentials or profile information as this is kept outside in an
external service. Unfortunately this decision also kept some prospects from
volunteering; a subject reported that he normally creates temporary email
accounts to try new web services. Other requirement was that they where all
adults, this age is 18 years in Mexico where the experiment was conducted. The
requirement of the social network account stopped several prospects from
volunteering, but on the other hand gave researchers access to public
information about the subjects. All the learners were either current students or
graduates of Computer Systems Engineering. Their ages were in the range of 18 to
30 years old. Although no experience in software programming was needed, as the
web tutorial's course is of a very basic level, all the participants were
required had completed at least one course in software programming. Results from
the initial survey are shown in Figure 4.
\begin{table*}[!t]
\centering
\caption{The performance of the classifiers.}
    \begin{tabular}{ | c | l | l | l | l | }
    \hline
    Affect          & Naïve Bayes           & J-48                & k-NN              & ANN \\
    \hline
    Flow/engaged    & 79.00\% (0.11) 0.50   & 80.48\% (0.10) 0.51 & 76.14(0.09) 0.43    & 78.43(0.12) 0.45 \\
    \hline
    Relaxation      & 71.10\% (0.09) 0.40   & 72.57\% (0.11) 0.45 & 69.14\%(0.08)  0.37 & 71.24\%(0.07)  0.34\\
    \hline
    Distraction     & 70.33\% (0.13)  0.43  & 71.00\% (0.10) 0.43 & 74.00\%(0.06)  0.50 & 72.52\%(0.07)  0.39\\
    \hline
    Frustration     & 74.71\% (0.09) 0.49   & 73.86\% (0.12) 0.45 & 72.62\%(0.09)  0.42 & 78.0\%(0.09)  0.51 \\
    \hline
    Boredom         & 74.71\% (0.13) 0.47   & 84.57\% (0.06) 0.59 & 83.81\%(0.06)  0.58 & 76.19\%(0.09)  0.45\\
    \hline
    \end{tabular}
\end{table*}

@@

The participants' interaction generated a total of 142 feature vectors one for each
successfully completed exercise. The affective states reported by learners after
completing each exercise was grouped in three classes: ‘Yes’, ‘Neutral’ and
‘No’. The class distribution of the emotions reported is shown in Figure 6. This
results show that the most common emotions were flow/engagement (72\%) and
relaxation (61\%) while few learners reported distraction (5\%), frustration
(8\%) and boredom (2\%). This distribution is different from what was reported
by D’Mello et al. \cite{bixler2013detecting} and Rodrigo et al. \cite{rodrigo2009affective}.
Although Flow/Engagement was
also the predominant class, the distribution is skewed to the first two. There
are some possible reasons for this; first the majority of students had
experience in programming, so learning new one is not that problematic, a second
reason could be the freedom users had to abandon the activities, perhaps
frustrated or bored learners simply quit the tutorial. A high number of learners
(49\%) did not completed any exercise and the once who completed more where also
the more experienced.

Classifiers where implemented using the software Rapid Miner using a feature
selection module as part of the process. A genetic algorithm was used in the
user selection module. In the end, the subset for the frustration classifier
consists of 11 features, and the subset for the boredom classifier consists of
13 features. The performance of the classifiers is shown in Table 1.
